A decision support tool for the Home Health Care Routing and Scheduling Problem

### 2022-08-28 tex/Paper/main.tex
new commit

\noindent Note that tardiness here refers only to the difference between the job start and its original time window. Jobs that have dependencies or require two nurses may be further restricted temporally. This can occur in practice, for example, in situations where a second dose of medication must be administered a few hours after the first dose. If two jobs $j_1$ and $j_2$ are dependent, this is modelled by imposing a second time window on $j_2$ that depends on the start time of $j_1$, $[t_{j_1} + e_{j_2}^{(j_1)}, t_{j_1} + l_{j_2}^{(j_1)}]$. Then, the same time window constraints hold:
\begin{subequations}
	\begin{align}
	t_{j_2} &\geq t_{j_1} + e^{(j_1)}_{j_2}, \label{eq:timewindow1}\\[3pt] % & \forall j \in J
	t_{j_2} &\leq t_{j_1} + l^{(j_1)}_{j_2}. \label{eq:timewindow2} % & \forall j \in J
	\end{align}
\end{subequations}

@@

\noindent Furthermore, we assume that waiting times are never larger than strictly necessary. To achieve this, we consider that the starting times of jobs are fixed to their earliest possible time, based on the order in which they appear in the solution. Note that this implies that the order in which nurse routes appear in a solution may have an impact on the solution quality. To mitigate this effect, in Section~\ref{sec:dst} we describe strategies to search different orderings of nurses when solving the problem.

@@

\begin{algorithm}[htbp!]
% \tiny
\caption{\textsc{CalculateJobTimes}$(i, r_i)$}
\label{alg:cjt}
\begin{algorithmic}[1]
	%\State feasible $\gets \False$
	\State Let $f$ be the initial shift for nurse $i$
	\State $a(i,k_{0}^i) \gets 0$
	\State $w_{ik_{0}^i} \gets 0$
	\State $s_{k_{0}^i} \gets 0$
%	\For{$j \in r_i$}
	\ForEach{job $j$ in nurse $i$'s route $r_i$}
		\State $a(i,j) \gets a(i, j-1) + w_{ij-1} + s_{j-1} + T_i(j-1, j)$
		\State $i' \gets \Null$, $j' \gets \Null$, DJ $\gets \False$, DS $\gets \False$ %, secondN $\gets \False$
		\If{$j \in J'$} % D : J' -> DJ, returns other job in pair with job j, NJ : J -> N, returns the nurse assigned to the job j
			\State $j' \gets D(j)$, $i' \gets NJ(j')$
			\If{$i$ appears later than $i'$ in $\mathcal{O}$}
				\If{$(j, j') \in DS$}
					\State DS $\gets \True$
				\Else \hspace{0.2mm} DJ $\gets \True$
				\EndIf
%				\State $[t_{j'} + e_j^{(j')}, t_{j'} + l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
				\State $[e_j^{(j')}, l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
			\EndIf
		\EndIf %if j in J'
		% 1. if nurse only has one shift
		\If{nurse $i$ only has one shift, $g_i = 1$}
			\State $w_{ij} \gets \max\{0, (e_j \chng{e_j^{(j')}} - a(i,j))\}$
			\State $t_j \gets a(i,j) + w_{ij}$
			\State $z_j \gets \max\{0, (t_j - l_j)\}$
			\If{DJ $= \True \Or$DS $= \True$}
				\State $\tilde{z}_j \gets \max\{0, (t_j - l_j^{(j')})\}$
			\EndIf
		% 2. Else nurse has multiple shifts, arrival is between shifts
		\ElsIf{$a(i,j) > v(\pi_{f}^{i}) \AAnd a(i,j) < u(\pi_{f+1}^{i}) \AAnd f \neq g_i$}
			\State $a(i,j)_{\textup{temp}}$ $\gets v(\pi_{f}^{i})$
			\State $t_j, w_{ij}, z_j, \tilde{z}_j, f \gets$ \textsc{FindValidTime}($i, j, f, a(i,j)_{\textup{temp}}, i', j'$, DS, DJ)
		% 3. Else arrival is in valid shift but job extends into break
		\ElsIf{$\max\{a(i,j), e_j\} + s_j > v(\pi_{f}^{i}) \AAnd f \neq g_i$}
			\State $t_j, w_{ij}, z_j, \tilde{z}_j, f \gets$ \textsc{FindValidTime}($i, j, f, a(i,j), i', j'$, DS, DJ)
		% 4. Job can be performed feasibly, or it is the final shift of nursei 
		\Else
			\State $w_{ij} \gets \max\{0, (e_j \chng{e_j^{(j')}} - a(i,j))\}$
			\State $t_j \gets a(i,j) + w_{ij}$
			\State $z_j \gets \max\{0, (t_j - l_j)\}$
			\If{DJ $= \True \Or$DS $= \True$}
				\State $\tilde{z}_j \gets \max\{0, (t_j - l_j^{(j')})\}$
			\EndIf
		\EndIf
	\EndFor
	\Return feasible schedule for nurse $i$
\end{algorithmic}
\end{algorithm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{algorithm}[htbp!]
% \scriptsize
\caption{\textsc{FindValidTime}($i, j, f, a(i,j), i', j'$, DS, DJ)}
\label{alg:fvt}
\begin{algorithmic}[1]
	\State feasible $\gets \False$
	\State $t_j \gets \Null$, $w_{ij} \gets \Null$, $z_j \gets \Null$, $\tilde{z}_j \gets \Null$
	\State $\tilde{f} \gets f$, $\tilde{a}(i,j) \gets a(i,j)$
	\If{DJ $= \True \Or$DS $= \True$}
%		\State $[t_{j'} + e_j^{(j')}, t_{j'} + l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
		\State $[e_j^{(j')}, l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
	\EndIf
	\While{$\Not$ feasible}
		\State $\tilde{f} \gets \tilde{f}+1$, $\tilde{a}(i,j) \gets u(\pi_{\tilde{f}}^{i})$
		% A. nurse i is the second nurse being assessed, i.e. job is DJ or DS.
%		\If{secondN $= \True$}
		\If{DJ $= \True \Or$DS $= \True$}
			% 1a. currentTime is LATER than latest allowed start time
%			\If{$\tilde{a}(i,j) > l_j^{(j')}$}
%				\If{$\tilde{a}(i,j) + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$}
%					\Continue
%				\Else
%					\State feasible $\gets \True$
%					\State $t_j \gets \tilde{a}(i,j)$
%					\State $w_{ij} \gets \max\{0, (t_j - a(i,j))\}$
%					\State $z_j \gets \max\{0, (t_j - l_j)\}$
%					\State $\tilde{z}_j \gets \max\{0, (t_j - l_j^{(j')})\}$
%					\State $f \gets \tilde{f}$
%				\EndIf
			% 2a. job cannot be performed feasibly in shift, extends into break
%			\ElsIf{$\max\{\tilde{a}(i,j), e_j^{(j')}\} + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$}
			\If{$\max\{\tilde{a}(i,j), e_j^{(j')}\} + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$} %\hspace{0.2mm}\textbf{continue}
				\Continue
			% 3a. nurse can perform job feasibly in shift OR shift \tilde{f} is the last shift of the day
			\Else
				\State feasible $\gets \True$
				\State $t_j \gets \tilde{a}(i,j)$
				\State $w_{ij} \gets \max\{0, (t_j - a(i,j))\}$
				\State $z_j \gets \max\{0, (t_j - l_j)\}$
				\State $\tilde{z}_j \gets \max\{0, (t_j - l_j^{(j')})\}$
				\State $f \gets \tilde{f}$
			\EndIf
		% End A. If secondN = True
		% B. nurse i is the first nurse assessing DJ/DS OR job is not a DJ/DS
		\Else
			% 1b. job cannot be performed feasibly in shift, extends into break
			\If{$\tilde{a}(i,j) + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$} %\hspace{0.2mm}\textbf{continue}
				\Continue
			% 2b. nurse can perform job feasibly in shift OR \tilde{f} is the last shift of the day
			\Else
				\State feasible $\gets \True$
				\State $t_j \gets \tilde{a}(i,j)$
				\State $w_{ij} \gets \max\{0, (t_j - a(i,j))\}$
				\State $z_j \gets \max\{0, (t_j - l_j)\}$
				\State $\tilde{z}_j \gets 0$
				\State $f \gets \tilde{f}$
			\EndIf
		\EndIf
	\EndWhile
	\Return starting time $t_j$, waiting time $w_{ij}$, tardiness $z_j$, gap tardiness $\tilde{z}_j$, and shift $f$
\end{algorithmic}
\end{algorithm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%This can occur in practice, for example, in situations where a second dose of medication must be administered a few hours after the first dose. If two jobs $j_1$ and $j_2$ are dependent, this is modelled by imposing a second time window on $j_2$ that depends on the start time of $j_1$, $[t_{j_1} + e_{j_2}^{(j_1)}, t_{j_1} + l_{j_2}^{(j_1)}]$. Then, the same time window constraints hold:
\noindent Note that tardiness here refers only to the difference between the job start and its original time window. Jobs that have dependencies or require two nurses may be further restricted temporally. This can occur in practice, for example, in situations where a set period of time is required between two separate doses of medication administered to a patient. Let $j_1$ and $j_2$ be two dependent jobs. Then, we model the additional temporal restriction by imposing a second time window of $j_2$ that depends on the start time of $j_1$, $[t_{j_1} + e_{j_2}^{(j_1)}, t_{j_1} + l_{j_2}^{(j_1)}]$, where $e_{j_2}^{(j_1)}$ and $l_{j_2}^{(j_1)}$ are the minimum and maximum time gaps from the start time of $j_1$ to the start time of $j_2$. Then, the same time window constraints hold:
\begin{subequations}
	\begin{align}
	t_{j_2} &\geq t_{j_1} + e^{(j_1)}_{j_2}, \label{eq:timewindow1}\\[3pt] % & \forall j \in J
	t_{j_2} &\leq t_{j_1} + l^{(j_1)}_{j_2}. \label{eq:timewindow2} % & \forall j \in J
	\end{align}
\end{subequations}

@@

\noindent Furthermore, we assume that waiting times are never larger than strictly necessary. To achieve this, we consider that the starting times of jobs are fixed to their earliest possible time, based on the order in which they appear in the solution. Note that for dependent jobs $(j_1, j_2) \in DJ$, the first nurse that is scheduled to do one of the two jobs sets the start time for that job, and therefore the other job acquires the additional time window. This implies that the order in which nurse routes appear in a solution may have an impact on the solution quality. To mitigate this effect, in Section~\ref{sec:dst} we describe strategies to search different orderings of nurses when solving the problem.

@@

\begin{algorithm}[htbp!]
% \tiny
\caption{\textsc{CalculateJobTimes}$(i, r_i)$}
\label{alg:cjt}
\begin{algorithmic}[1]
	%\State feasible $\gets \False$
	\State Let $f$ be the initial shift for nurse $i$
	\State $a(i,k_{0}^i) \gets 0$
	\State $w_{ik_{0}^i} \gets 0$
	\State $s_{k_{0}^i} \gets 0$
%	\For{$j \in r_i$}
	\ForEach{job $j$ in nurse $i$'s route $r_i$}
		\State $a(i,j) \gets a(i, j-1) + w_{ij-1} + s_{j-1} + T_i(j-1, j)$
		\State $i' \gets \Null$, $j' \gets \Null$, DJ $\gets \False$, DS $\gets \False$ %, secondN $\gets \False$
		\If{$j \in J'$} % D : J' -> DJ, returns other job in pair with job j, NJ : J -> N, returns the nurse assigned to the job j
			\State $j' \gets D(j)$, $i' \gets NJ(j')$
			\If{$i$ appears later than $i'$ in $\mathcal{O}$}
				\If{$(j, j') \in DS$}
					\State DS $\gets \True$
				\Else \hspace{0.2mm} DJ $\gets \True$
				\EndIf
				\State $[t_{j'} + e_j^{(j')}, t_{j'} + l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
%				\State $[e_j^{(j')}, l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
			\EndIf
		\EndIf %if j in J'
		% 1. if nurse only has one shift
		\If{nurse $i$ only has one shift, $g_i = 1$}
			\If{DJ $= \True \Or$DS $= \True$}			
				\State $w_{ij} \gets \max\{0, ((t_{j'} + e_j^{(j')}) - a(i,j))\}$
				\State $t_j \gets a(i,j) + w_{ij}$
				\State $\tilde{z}_j \gets \max\{0, (t_j - (t_{j'} + l_j^{(j')}))\}$
			\Else
				\State $w_{ij} \gets \max\{0, (e_j - a(i,j))\}$
				\State $t_j \gets a(i,j) + w_{ij}$
			\EndIf
			\State $z_j \gets \max\{0, (t_j - l_j)\}$
		% 2. Else nurse has multiple shifts, arrival is between shifts
		\ElsIf{$a(i,j) > v(\pi_{f}^{i}) \AAnd a(i,j) < u(\pi_{f+1}^{i}) \AAnd f \neq g_i$}
			\State $a(i,j)_{\textup{temp}}$ $\gets v(\pi_{f}^{i})$
			\State $t_j, w_{ij}, z_j, \tilde{z}_j, f \gets$ \textsc{FindValidTime}($i, j, f, a(i,j)_{\textup{temp}}, i', j'$, DS, DJ)
		% 3. Else arrival is in valid shift but job extends into break
		\ElsIf{$\max\{a(i,j), e_j\} + s_j > v(\pi_{f}^{i}) \AAnd f \neq g_i$}
			\State $t_j, w_{ij}, z_j, \tilde{z}_j, f \gets$ \textsc{FindValidTime}($i, j, f, a(i,j), i', j'$, DS, DJ)
		% 4. Job can be performed feasibly, or it is the final shift of nursei 
		\Else
			\If{DJ $= \True \Or$DS $= \True$}			
				\State $w_{ij} \gets \max\{0, ((t_{j'} + e_j^{(j')}) - a(i,j))\}$
				\State $t_j \gets a(i,j) + w_{ij}$
				\State $\tilde{z}_j \gets \max\{0, (t_j - (t_{j'} + l_j^{(j')}))\}$
			\Else
				\State $w_{ij} \gets \max\{0, (e_j - a(i,j))\}$
				\State $t_j \gets a(i,j) + w_{ij}$
			\EndIf
			\State $z_j \gets \max\{0, (t_j - l_j)\}$
		\EndIf
	\EndFor
	\Return feasible schedule for nurse $i$
\end{algorithmic}
\end{algorithm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{algorithm}[htbp!]
% \scriptsize
\caption{\textsc{FindValidTime}($i, j, f, a(i,j), i', j'$, DS, DJ)}
\label{alg:fvt}
\begin{algorithmic}[1]
	\State feasible $\gets \False$
	\State $t_j \gets \Null$, $w_{ij} \gets \Null$, $z_j \gets \Null$, $\tilde{z}_j \gets \Null$
	\State $\tilde{f} \gets f$, $\tilde{a}(i,j) \gets a(i,j)$
	\If{DJ $= \True \Or$DS $= \True$}
		\State $[t_{j'} + e_j^{(j')}, t_{j'} + l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
%		\State $[e_j^{(j')}, l_j^{(j')}] \gets$ dependent time window, based on $t_{j'}$ set by $i'$
	\EndIf
	\While{$\Not$ feasible}
		\State $\tilde{f} \gets \tilde{f}+1$, $\tilde{a}(i,j) \gets u(\pi_{\tilde{f}}^{i})$
		% A. nurse i is the second nurse being assessed, i.e. job is DJ or DS.
%		\If{secondN $= \True$}
		\If{DJ $= \True \Or$DS $= \True$}
			% 1a. currentTime is LATER than latest allowed start time
%			\If{$\tilde{a}(i,j) > l_j^{(j')}$}
%				\If{$\tilde{a}(i,j) + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$}
%					\Continue
%				\Else
%					\State feasible $\gets \True$
%					\State $t_j \gets \tilde{a}(i,j)$
%					\State $w_{ij} \gets \max\{0, (t_j - a(i,j))\}$
%					\State $z_j \gets \max\{0, (t_j - l_j)\}$
%					\State $\tilde{z}_j \gets \max\{0, (t_j - l_j^{(j')})\}$
%					\State $f \gets \tilde{f}$
%				\EndIf
			% 2a. job cannot be performed feasibly in shift, extends into break
%			\ElsIf{$\max\{\tilde{a}(i,j), e_j^{(j')}\} + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$}
%			\If{$\max\{\tilde{a}(i,j), e_j^{(j')}\} + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$} %\hspace{0.2mm}\textbf{continue}
			\If{$\max\{\tilde{a}(i,j), (t_{j'} + e_j^{(j')})\} + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$} %\hspace{0.2mm}\textbf{continue}
				\Continue
			% 3a. nurse can perform job feasibly in shift OR shift \tilde{f} is the last shift of the day
			\Else
				\State feasible $\gets \True$
				\State $t_j \gets \tilde{a}(i,j)$
				\State $w_{ij} \gets \max\{0, (t_j - a(i,j))\}$
				\State $z_j \gets \max\{0, (t_j - l_j)\}$
				\State $\tilde{z}_j \gets \max\{0, (t_j - (t_{j'} + l_j^{(j')}))\}$
				\State $f \gets \tilde{f}$
			\EndIf
		% End A. If secondN = True
		% B. nurse i is the first nurse assessing DJ/DS OR job is not a DJ/DS
		\Else
			% 1b. job cannot be performed feasibly in shift, extends into break
			\If{$\tilde{a}(i,j) + s_j > v(\pi_{\tilde{f}}^{i}) \AAnd \tilde{f} \neq g_i$} %\hspace{0.2mm}\textbf{continue}
				\Continue
			% 2b. nurse can perform job feasibly in shift OR \tilde{f} is the last shift of the day
			\Else
				\State feasible $\gets \True$
				\State $t_j \gets \tilde{a}(i,j)$
				\State $w_{ij} \gets \max\{0, (t_j - a(i,j))\}$
				\State $z_j \gets \max\{0, (t_j - l_j)\}$
				\State $\tilde{z}_j \gets 0$
				\State $f \gets \tilde{f}$
			\EndIf
		\EndIf
	\EndWhile
	\Return starting time $t_j$, waiting time $w_{ij}$, tardiness $z_j$, gap tardiness $\tilde{z}_j$, and shift $f$
\end{algorithmic}
\end{algorithm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
